---
title: "Corporate Financial Resilience During COVID-19: Sweden vs Cayman Islands"
author: "AWADHESKUMAR THAKKAR"
date: "Oct 6, 2025"
format:
    html: 
        output-file: financial_analysis.html
        css: "assignment.css"
        embed-resources: true
        code-fold: true
        toc: true
        
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
# Set up chunk for all slides
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r}
#| echo: false
# Load libraries
library(tidyverse)
library(conflicted)
library(here)
library(dplyr)
library(visdat)
library(tidyverse)
library(tourr)

library(ggplot2)
library(skimr)       # tidy summaries
library(kableExtra)  # neat tables
theme_set(theme_minimal())


conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)

# Plot options and themes
options(
  digits = 2,
  width = 60,
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)

theme_set(theme_bw(base_size = 14) +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title.position = "plot",
     plot.title = element_text(size = 24),
     panel.background = element_rect(fill = 'transparent', colour = NA),
     legend.background = element_rect(fill = 'transparent', colour = NA),
     legend.key = element_rect(fill = 'transparent', colour = NA)
   )
)
```

```{r load}
# Load each year’s file
load(here("osiris", "osiris_Sweden_2018.rda"))
load(here("osiris", "osiris_Sweden_2019.rda"))
load(here("osiris", "osiris_Sweden_2020.rda"))
load(here("osiris", "osiris_Sweden_2021.rda"))
```

# Introduction

This project explores how firms responded to the economic shock of COVID-19 by analysing financial data from Sweden and the Cayman Islands between 2017 and 2021.

The analysis focuses on revenue, profitability, and sector-level trends to understand resilience patterns across two contrasting economic systems.

Sweden is a diverse and highly competitive economy, mainly focused on
goods and manufactures with a growth in service sectors such as
information and communications technology (3). According to the OCDE
Sweden's economy is characterised by highly educated workforce, strong
institutions, low income inequality and an extensive welfare system that
supports productivity and helps individuals participate effectively in
the labor market (4). Some of the most famous Swedish companies around
the world are Volvo, Ikea, H&M and Spotify (5)

According to the World Bank, the COVID-19 pandemic between 2020 and 2022
triggered the largest global economic crisis of this century (6\*). With
this report, we aim to investigate how the pandemic impacted the
financial performance of Swedish companies. To achieve this, we will use
data from OSIRIS database, which provides standardised financial
statements of public and private companies worldwide (2). Our analysis
will focus on key financial indicators such as liabilities and debt,
revenue, total assets, earnings before and after tax, revenues, profit
margins and employment. By examining changes in these indicators, we
provide an integral analysis of how Sweden's Corporate sector responded
to this crisis and observe any recovery dynamics that might have emerged
in the post-pandemic period.

# Data description

The dataset covers the period 2017 to 2021, allowing us to compare
betwen the pre-pandemic (2017-2019) and pandemic periods (2020 and
2021). The data contains 240 financial variables and around 1800 records
per year. The combined dataset includes company identifiers (NAICS
classification) and key financial indicators per year.

To ensure the validity of our analysis, we remove duplicated records
using firm and year identifiers, variable names are standardised accross
years. Outliers will be evaluated through visual inspections (box plots)
to detect any potential data entry or reporting anomaly. To ensure that
our analysis only focuses on reliable data, we exclude any variable that
is considered irrelevant, and retain only variables that contain less
than 60% of missing values.

```{r structure}
#checking the structure of the datasets
dim(osiris_2018)
dim(osiris_2019)
dim(osiris_2020)
dim(osiris_2021)
unique(osiris_2018$year)
unique(osiris_2019$year)
unique(osiris_2020$year)
unique(osiris_2021$year)
```

```{r}
names(osiris_2018)
```

```{r}
# combining the datasets and ensure no duplications
sweden_all <- rbind(osiris_2018, osiris_2019, osiris_2020, osiris_2021) %>% distinct()

unique(sweden_all$`NAICS Core Code (caics12cod)`)

sweden_all <- sweden_all %>%
  rename(naics = `NAICS Core Code (caics12cod)`)

range(sweden_all$ naics, na.rm = TRUE)
```

```{r}
vis_dat(osiris_2018)
dim(sweden_all)          # total rows and columns
unique(osiris_2018$year)   # how many rows per year
head(sweden_all)
```

```{r check percentages of na and drop some useless columns}
na_per <- sweden_all %>% 
            is.na() %>% 
            colMeans() %>% 
            sort(decreasing = TRUE)

# contact information is also not useful to keep
# also dont need country information since all are known to be in sweden
useless_cols <- c('Email Address (email)', 
                  'Web Site Address (www)',
                  'FAX - Main Fax Number (fax)',
                  'PHONE - Main Phone Number (phone)',
                  'Country (country)',
                  'Country Code (cntrycde)'
                  )

na_keep <- na_per[na_per < 0.6] %>% names() %>% 
            setdiff(useless_cols)

sweden_dropped <- sweden_all %>% select(all_of(na_keep))
```

#### grouping the naics into different sectors from the NAICS website

```{r}
sweden_all <- sweden_all %>%
  mutate(naics_sector = substr(naics, 1, 2))
sector_labels <- c(
  "11" = "Agriculture, Forestry, Fishing & Hunting",
  "21" = "Mining, Quarrying, Oil & Gas Extraction",
  "22" = "Utilities",
  "23" = "Construction",
  "31" = "Manufacturing",
  "32" = "Manufacturing",
  "33" = "Manufacturing",
  "42" = "Wholesale Trade",
  "44" = "Retail Trade",
  "45" = "Retail Trade",
  "48" = "Transportation & Warehousing",
  "49" = "Transportation & Warehousing",
  "51" = "Information",
  "52" = "Finance & Insurance",
  "53" = "Real Estate & Rental",
  "54" = "Professional, Scientific & Technical Services",
  "55" = "Management of Companies",
  "56" = "Administrative & Support",
  "61" = "Education Services",
  "62" = "Health Care & Social Assistance",
  "71" = "Arts, Entertainment & Recreation",
  "72" = "Accommodation & Food Services",
  "81" = "Other Services",
  "92" = "Public Administration"
)

sweden_all <- sweden_all %>%
  mutate(sector_name = sector_labels[naics_sector])

```

# IDA

### Sample and coverage overview

```{r companies-per-year, message=FALSE, warning=FALSE}
# Uses sweden_all already created above
yr_counts <- sweden_all |>
  dplyr::rename(firm_id = `BvD ID Number (os_id_number)`) |>
  dplyr::summarise(
    n_records   = dplyr::n(),
    n_companies = dplyr::n_distinct(firm_id),
    .by = year
  ) |>
  dplyr::arrange(year)

kableExtra::kbl(yr_counts, caption = "Records & unique companies by year") |>
  kableExtra::kable_styling(full_width = FALSE)

```

From 2017 to 2021, record counts remained quite consistent, between 825
to 933 per year. Only a few companies have more than one record in a
single year since unique-company counts track records closely (gaps of
only approx 5–8). Instead of a collapse in coverage, the small rise
through 2019 and following slight decline in 2020–2021 are consistent
with the slight disruption in reporting throughout the COVID period.

```{r companies-by-year-sector, message=FALSE, warning=FALSE}
# Prefer sector_name, fall back to naics_sector
sector_col <- dplyr::case_when(
  "sector_name"  %in% names(sweden_all) ~ "sector_name",
  "naics_sector" %in% names(sweden_all) ~ "naics_sector",
  TRUE ~ NA_character_
)
stopifnot(!is.na(sector_col))

yr_sector_counts <- sweden_all |>
  dplyr::rename(firm_id = `BvD ID Number (os_id_number)`) |>
  dplyr::mutate(sector = tidyr::replace_na(.data[[sector_col]], "Unknown")) |>
  dplyr::summarise(
    n_records   = dplyr::n(),
    n_companies = dplyr::n_distinct(firm_id),
    .by = c(year, sector)
  ) |>
  dplyr::arrange(year, dplyr::desc(n_records))

yr_sector_distinct_counts <- yr_sector_counts  |>
  select(-n_records) |>
  pivot_wider(names_from = year,
              values_from = n_companies)

kableExtra::kbl(yr_sector_distinct_counts, caption = "Records & unique companies by year × sector") |>
  kableExtra::kable_styling(full_width = FALSE)

```

Every year, manufacturing (about 380 businesses) dominates the sample,
followed by professional, scientific and technical services, wholesale
trade, real estate and rental, and information. Throughout,
pandemic-sensitive industries like lodging and food services continue to
hold a relatively tiny percentage.

From 2017 to 2021, the sector composition remained surprisingly
consistent. Would be ideal to use of sector-stratified views thereafter
to avoid the effects of big manufacturing numbers masking smaller sector
changes.

#### Choosing a few important variables to start with. these are subjective. The analysis flow is taken from week 3 slides.

```{r}
# IDA: Sweden financial data

# Select relevant variables
key_vars <- c("Total Assets (data13077)",
              "Total Liabilities and Debt (data14022)",
              "Total revenues (data13004)",
              "Gross sales (data13000)",
              "Earnings before tax (data13034)",
              "Earnings after tax (data13037)",
              "Profit Margin (%) (data31005)",
              "No. Of employed (data23000)",
              "naics_sector",
              "sector_name",
              "year")

sweden_ida <- sweden_all %>% select(any_of(key_vars))

# 1. Structure & types

# 2. Numeric summary
summary(sweden_ida)

# 3. Missingness
round(colMeans(is.na(sweden_ida))*100, 1)

# 4. Plausibility checks
range(sweden_ida$`Profit Margin (%) (data31005)`, na.rm = TRUE)

```

**Simplifying variables names**

```{r}

# --- Rename core variables for clarity 
sweden_ida_sub <- sweden_ida %>%
  rename(
    total_assets       = `Total Assets (data13077)`,
    total_liabilities  = `Total Liabilities and Debt (data14022)`,
    total_revenues     = `Total revenues (data13004)`,
    gross_sales        = `Gross sales (data13000)`,
    earnings_before_tax= `Earnings before tax (data13034)`,
    earnings_after_tax = `Earnings after tax (data13037)`,
    profit_margin      = `Profit Margin (%) (data31005)`,
    employees          = `No. Of employed (data23000)`
  )

```

#### Patterns within total assets and liabilities and debt

Here, some basic investigation into the patterns within the total assets
and liabilities is done. First, a simple histogram to see the
distribution of assets and liabilities in a single year to get an idea
of the distribution.

```{r investigating assets and liabilities - graphical}
# initial investigation will not use a tsibble, only looking at means across all observations in the 5 years
sweden_al_sub <- sweden_ida_sub %>% 
      select(year, total_assets, total_liabilities, sector_name) %>% 
          mutate(l_assets = log10(total_assets + 1),
                 l_liab = log10(total_liabilities+ 1))
# incorporated log statistics based on below observations for easier readability
```

Separating the data, we now do a simple plot to see if there's anything
obvious, doing a simple histogram (Lecture 1, slide 34).

```{r plot histograms}
# first look at one year to see the distribution
sweden_al_sub %>% filter(year == 2019) %>% 
  ggplot(aes(x = total_assets)) + 
  geom_histogram(bins = 60) + 
  labs(x = "Total Assets")
```

We can see severe right skewness (which makes sense for financial data),
so we will apply a log scale (Lecture 5, slide 28) to make the plot
easier to understand and visualise.

```{r plot histograms skew}
# first look at one year to see the distribution
sweden_al_sub %>% filter(year == 2019) %>% 
  ggplot(aes(x = total_assets)) + 
  geom_histogram(bins = 60) +
  scale_x_log10() + 
  labs(x = "Total Assets")
```

# EDA (with some analysis between)

Looking at this plot, we see a better distribution of the data, where it
is relatively normally distributed, so we can use this to generate our
means and medians to get a better idea of the distributions (scaling
with log and then transforming back) (Lecture 5, slide 29). We also need
to add 1 when transforming with the log to ensure we don't have negative
infinity skewing our means. (This was moved a couple chunks up since
this will be reused).

```{r assets + liab - numerical }
sweden_al_summary <- sweden_al_sub  %>% 
          group_by(year) %>% 
          summarise(`Mean Assets` = 10^mean(l_assets, na.rm = TRUE) - 1,
                    `Median Assets` = 10^median(l_assets, na.rm = TRUE) - 1,
                    `Mean Liabilities` = 10^mean(l_liab, na.rm = TRUE) - 1,
                    `Median Liabilities` = 10^median(l_liab, na.rm = TRUE) - 1)
sweden_al_summary
```

We see that there is a general increasing trend, although interestingly
there appears to be a dip in 2019, although this may be a pattern of the
collection methods. To get a better understanding of the distribution,
we can use a boxplot to visualise this (Lecture 5, slide 17).

```{r assets + libs - boxplot}
sweden_al_sub %>% 
  ggplot(aes(x = factor(year), y = l_assets)) +
  geom_boxplot() +
  labs(
    x = "Year",
    y = "Total assets (Logarithmic)"
  )

sweden_al_sub %>%
  ggplot(aes(x = factor(year), y = l_liab)) +
  geom_boxplot() +
  labs(
    x = "Year",
    y = "Total liabilities and debt (Logarithmic)"
  )
```

We can see that overall, there is no major shifts across the years,
however interestingly there is still an increasing pattern across all
the years for both total assets and liabilities, but not a significant
shift overall.

To get a better idea of underlying patterns, we can do some stratified
visualisation based on the sectors. (Lecture 6, Slide 19)

```{r scatterplot facet, fig.width = 15, fig.height = 6}
sweden_al_sub %>% 
  ggplot(aes(x = factor(year), y = l_assets)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~sector_name, ncol = 10, labeller = label_wrap_gen(width = 20)) +
  theme(aspect.ratio = 1) +
  labs(
    x = "Year",
    y = "Total Assets (Logarithmic)"
  )

sweden_al_sub %>% 
  ggplot(aes(x = factor(year), y = l_liab)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~sector_name, ncol = 10, labeller = label_wrap_gen(width = 20)) +
  theme(aspect.ratio = 1) +
  labs(
    x = "Year",
    y = "Total Liabilities and Debt (Logarithmic)"
  )
```

Looking at this distribution across the different sectors, we can notice
there are numerous more interesting changes across the years for many
sectors. For example, there are significant spikes in pandemic years
(2020 and 2021) in the Accommodation & Food Services and Public
Administration. When comparing this to the number of companies across
the years, the increase in total assets for Accommodation & Food
Services is likely attributed to the poorer performing companies having
to shutdown, and only high performing companies remaining. Conversely,
the pattern shown in Public Administration is likely due to the
relatively large increase in the number of companies in the sector in
the pandemic years.

Looking at larger sectors, such as Manufacturing, and Professional,
Scientific & Technical services, and Wholesale Trade, there are no
significant shifts, although they do have slightly upward trends which
aligns with previous inspections of general increases across each year,
with no noticeable impact on the total assets or liabilities.

Sectors where this pattern noticeably differs is within the Arts,
Entertainment & Recreation, likely due to overall reduced recreational
usage, as well as minor drops within the Agriculture, Forestry, Fishing
& Hunting and Transportation & Warehousing sectors, which may be
attributed to limited outdoor access during pandemic control measures.
We also see a noticeable shift in the median assets and liabilities of
Education Services, perhaps a result of government or external
investment in resources required for remote learning capabilities during
the pandemic.

#### Earnings before & after tax

**Investigation into the patterns in earnings before tax and after tax
is done below**

```{r}
ebt <- sweden_ida_sub %>% 
  select(year,earnings_before_tax, earnings_after_tax)

#checking if major na values are present.

ebt %>% 
  filter(is.na(earnings_after_tax)) %>% 
  count()
```

**doing a simple histogram to check distribution. from lecture 1**

```{r}
ggplot(ebt, aes(x=earnings_before_tax))+
  geom_histogram(bins=60, fill="pink")

ggplot(ebt, aes(x=earnings_after_tax))+
  geom_histogram(bins = 60, fill = "green")
```

```{r}
ebt %>%
  group_by(year) %>%
  summarise(
    min_value    = min(earnings_before_tax, na.rm = TRUE),
    median_value = median(earnings_before_tax, na.rm = TRUE),
    mean_value   = mean(earnings_before_tax, na.rm = TRUE),
    max_value    = max(earnings_before_tax, na.rm = TRUE)
  )

```

Mean positive + median negative possible explains strong right skew. we
can validate that with the visualisations done before. A few firms
report very high profits.

**Log transforming the earnings to better visualise. (from week 5 slide
28)**

```{r}
ggplot(ebt, aes(x=abs(earnings_before_tax)))+
  geom_histogram(bins=60, fill="pink")+
  scale_x_log10()+
  facet_wrap(~year)

```

Creating the yearly summary for log(absolute earnings)

```{r}
ebt_year_summary <- ebt %>%
  mutate(log_abs_ebt = log10(abs(earnings_before_tax) + 1)) %>%  # add +1 to avoid log(0)
  group_by(year) %>%
  summarise(
    mean_log_ebt   = mean(log_abs_ebt, na.rm = TRUE),
    median_log_ebt = median(log_abs_ebt, na.rm = TRUE)
  )
ebt_year_summary

```

To avoid dropping firms with negative earnings, we have used absolute
values of earnings. This allows analysis of earnings magnitudes. the
values given above are exponents to base 10.

The proportion of profit- and loss-making firms remained roughly
constant over 2017–2021, suggesting no major structural shift in
profitability. However, the log-transformed absolute earnings before tax
increased sharply in 2021, indicating that the size of both profits and
losses expanded. This pattern likely reflects a bit of volatility during
the pandemic, where some firms experienced exceptional gains while
others suffered larger losses.

#### Classify earnings in the direction of either profit or loss

```{r}
# 1) Tag direction using BEFORE-tax earnings
ebt_direction <- ebt %>%
  mutate(
    direction = case_when(
      earnings_before_tax > 0 ~ "Profit",
      earnings_before_tax < 0 ~ "Loss",
      TRUE ~ "Zero"
    )
  )
```

**Checking profit vs loss firms by year**

```{r}
# Table
ebt_direction %>%
  group_by(year, direction) %>%
  summarise(
    n_firms = n(),
    median_ebt = median(earnings_before_tax, na.rm = TRUE),
    mean_ebt   = mean(earnings_before_tax,   na.rm = TRUE),
    .groups = "drop"
  )

```

Median profts and median losses remain fairly stable through 2017-2020.
However both profits and losses increase in magnitude in 2021. The high
profits in 2021 c

**plotting it using barplots**

```{r}
# Plot (counts)
ebt_direction %>%
  group_by(year, direction) %>%
  summarise(n_firms = n(), .groups = "drop") %>%
  ggplot(aes(x = year, y = n_firms, fill = direction)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of Profit vs Loss Firms by Year",
    x = "Year", y = "Number of Firms", fill = "Earnings Direction"
  ) +
  theme_minimal()

```

Not much of a difference between numbers of profit and loss amking
firms. the ratios remain balanced throughout the years.

Now, Using the existing earnings direction categories, we shall compare
the median magnitudes of before and after-tax earnings on a log10 scale

```{r}
# Add after-tax variables and log transforms
ebt_compare <- ebt_direction %>%
  mutate(
    log_abs_ebt = log10(abs(earnings_before_tax) + 1),
    log_abs_eat = log10(abs(earnings_after_tax)  + 1)
  )

# Summarise by year and direction
summary_simple <- ebt_compare %>%
  group_by(year, direction) %>%
  summarise(
    median_before = median(log_abs_ebt, na.rm = TRUE),
    median_after  = median(log_abs_eat, na.rm = TRUE),
    .groups = "drop"
  )

# Plot, excluding Zero direction
summary_simple %>%
  filter(direction %in% c("Profit", "Loss")) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = median_before, color = "Before Tax", group = direction), linewidth = 1) +
  geom_line(aes(y = median_after,  color = "After Tax",  group = direction),
            linewidth = 1, linetype = "dashed") +
  facet_wrap(~ direction, nrow = 1) +
  labs(
    title = "Median log10(|Earnings|) Before vs After Tax by Year",
    x = "Year", y = "Median log10(|Earnings| + 1)",
    color = "Measure"
  ) +
  theme_minimal()


```

The after-tax line is slightly above the before-tax. profits and losses
increased overall in 2021 with greater volatility. some reasons could be
post-pandemic recovery, and corporate tax rate cuts in Sweden in the
year 2021.

During COVID-19 some companies globally decided to downsize their
workforce, then we can explore the number of employees as a proxy of
financial health. However in the following summary, we observe that in a
pre-pandemic period the average number of employees was lower than in
the pandemic period. It is an interesting pattern that should be
investigated, based on Lecture 3 slides (slide 46) we can run a t-test
to do hypothesis testing.

```{r}
# initial investigation will not use a tsibble, only looking at means across all observations in the 5 years
sweden_small <- sweden_ida_sub %>% 
      select(year, employees, profit_margin, naics_sector)

sweden_small <- sweden_small %>% 
  mutate(pandemic_period = case_when(
    year < 2020 ~ 'pre-pandemic',
    year %in% c (2020, 2021, 2022) ~ 'pandemic',
    year >= 2023 ~ 'post-pandemic'
  ))

summary_by_period <- sweden_small %>%
  group_by(pandemic_period) %>%
  summarise(
    avg_profit_margin = mean(profit_margin, na.rm = TRUE),
    avg_employees = mean(employees, na.rm = TRUE),
    n_firms = n()
  )

summary_by_period

#week3 slide 46 

tt <- with(sweden_small,
  t.test(
    employees[pandemic_period == "pre-pandemic"],
    employees[pandemic_period == "pandemic"],
    paired = FALSE
  )
)

tt

```

Based on the results of the t-test we find that there is no significant
difference in the means of the number of employees per company between
the pandemic and postpandemic periods (p-value =1). This result is an
interesting finding as it contradicts the expectation that as a result
of lock downs (specially for sectors like mining or manufacture)
lockdowns pressured companies to downsize (3). This result can indicate
that the trend in Nordic companies, like the ones included in this
dataset are resilient or that Sweeden's pandemic strategy helped
companies to retain staff and can lead to further investigation (2).

Based on Week2, we produce scaled box-plots (y scale log 10) to compare
the number of employees in the biggest sectors (21 = "Mining, Quarrying,
Oil & Gas Extraction", 32 and "33 = "Manufacturing", 42 = "Wholesale
Trade", 51 = "Information", 52 = "Finance & Insurance", 53 = "Real
Estate & Rental", 54 = "Professional, Scientific & Technical Services",
56 = "Administrative & Support, 62 = "Health Care & Social Assistance").

```{r}

sweden_plot <- sweden_small %>%
  filter(!is.na(employees), !is.na(naics_sector), !is.na(pandemic_period))

# Filter to sectors with at least 30 firms
top_sectors <- sweden_plot %>%
  group_by(naics_sector) %>%
  tally() %>%
  filter(n >= 100) %>%
  pull(naics_sector)

sweden_plot <- sweden_plot %>%
  filter(naics_sector %in% top_sectors)

ggplot(sweden_plot, aes(x = naics_sector, y = employees)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4) +
  scale_y_log10()+
  labs(title = "Number of Employees per Sector",
       x = "NAICS Sector",
       y = "Employees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

Now we will look into the patterns between the two variables Total
Revenues and Gross sales. Before commencing, we will. take a look at the
data and create a quick profile for the 2 variables.

```{r}
# a small focus frame
vars_use <- sweden_all |>
  dplyr::select(
    year,
    total_revenues = `Total revenues (data13004)`,
    gross_sales    = `Gross sales (data13000)`
  ) |>
  mutate(
    total_revenues = as.numeric(total_revenues),
    gross_sales    = as.numeric(gross_sales)
  )

# an overall summary (types, missing, basic stats)
skimr::skim(vars_use[, c("total_revenues","gross_sales")])

# checking missingness by year
vars_use |>
  summarise(
    n = dplyr::n(),
    missing_total_revenues = sum(is.na(total_revenues)),
    missing_gross_sales    = sum(is.na(gross_sales)),
    .by = year
  ) |>
  arrange(year) |>
  kableExtra::kbl(caption = "Missingness by year: total_revenues & gross_sales") |>
  kableExtra::kable_styling(full_width = FALSE)

# scale check medians & IQR by year (robust, handles heavy tails)
vars_use |>
  summarise(
    n = dplyr::n(),
    median_total_revenues = median(total_revenues, na.rm = TRUE),
    IQR_total_revenues    = IQR(total_revenues, na.rm = TRUE),
    median_gross_sales    = median(gross_sales, na.rm = TRUE),
    IQR_gross_sales       = IQR(gross_sales, na.rm = TRUE),
    .by = year
  ) |>
  arrange(year) |>
  kableExtra::kbl(caption = "Robust summaries by year (median & IQR)") |>
  kableExtra::kable_styling(full_width = FALSE)

# zeros/negatives (checkin in case of potential log scales later)
vars_use |>
  summarise(
    zeros_total_revenues = sum(!is.na(total_revenues) & total_revenues == 0),
    negatives_total_revenues = sum(!is.na(total_revenues) & total_revenues < 0),
    zeros_gross_sales = sum(!is.na(gross_sales) & gross_sales == 0),
    negatives_gross_sales = sum(!is.na(gross_sales) & gross_sales < 0)
  ) |>
  kableExtra::kbl(caption = "Zero/negative counts") |>
  kableExtra::kable_styling(full_width = FALSE)

```

Will look at yearly medians with IQR ribbons which uses dplyr + ggplot2;
Week 2 & Week 5 content:

```{r}
q_by_year <- vars_use |>
  tidyr::pivot_longer(c(total_revenues, gross_sales),
                      names_to="metric", values_to="value") |>
  dplyr::summarise(
    n = dplyr::n(),
    q25 = quantile(value, .25, na.rm=TRUE),
    med = median(value, na.rm=TRUE),
    q75 = quantile(value, .75, na.rm=TRUE),
    .by = c(year, metric)
  )

ggplot(q_by_year, aes(x = year, y = med)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.25) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_log10() +
  facet_wrap(~ metric, scales = "free_y") +
  labs(title = "Median and IQR by year (log y)",
       y = "Value (log10)")
```

For both gross_sales and total_revenues, the median + IQR (log-y) plot
displays a gradual increase from 2017 to 2019, a slight decline in 2020,
and a return in 2021, exactly the "pandemic dent then recovery" pattern
we anticipated. The broad, rather stable IQR ribbons show disturbance
without a structural collapse, indicating heavy-tailed but generally
stable dispersion across years.

Now will look at indexed medians which employs dplyr/ggplot2 from Weeks
2 (wrangling) and 5 (summary plots) to compare trends directly on the
same scale.

```{r}
# Median by year & metric with no residual grouping
medyr <- sweden_all |>
  transmute(
    year,
    total_revenues = as.numeric(`Total revenues (data13004)`),
    gross_sales    = as.numeric(`Gross sales (data13000)`)
  ) |>
  pivot_longer(c(total_revenues, gross_sales),
               names_to = "metric", values_to = "value") |>
  group_by(year, metric) |>
  summarise(med = median(value, na.rm = TRUE), .groups = "drop")

# Pulling 2019 baselines per metric
base19 <- medyr |>
  filter(year == 2019) |>
  rename(baseline = med) |>
  select(metric, baseline)

# Joining baseline and computing index (falls back to first non-missing if 2019 is NA/missing)
med_idx <- medyr |>
  left_join(base19, by = "metric") |>
  group_by(metric) |>
  mutate(
    baseline = ifelse(is.na(baseline),
                      med[which.min(abs(year - 2019))],  # fallback baseline
                      baseline),
    idx_2019 = 100 * med / baseline
  ) |>
  ungroup()

ggplot(med_idx, aes(x = year, y = idx_2019, colour = metric)) +
  geom_hline(yintercept = 100, linetype = 2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Indexed medians (2019 = 100)",
       y = "Median index", x = "Year", colour = NULL)
```

Gross sales and total revenues move nearly in unison, with a distinct
decline in 2020 (\~93–94) and a significant rebound in 2021 (\~111–112),
according to the indexed-median plot (2019 = 100). The dashed line at
100 serves as a marker for changes in relation to the pre-pandemic
baseline, and values are based on raw medians with na.rm=TRUE.

We will now compare gross sales to total revenues as a ratio by year to
verify consistency between the two metrics. By plotting
log10(gross_sales/total_revenues), we may determine if typical firms sit
close to equality and how dispersion varies over time. This places
"parity" at 0 (i.e., 1×). Thisemploys distributions with modified scales
in ggplot2 (Week 5–6).

```{r}
ratio_df <- sweden_all |>
transmute(
year,
total_revenues = as.numeric(`Total revenues (data13004)`),
gross_sales    = as.numeric(`Gross sales (data13000)`)
) |>

# keep comparable values
filter(!is.na(total_revenues), !is.na(gross_sales),
total_revenues > 0, gross_sales >= 0) |>
mutate(
ratio      = gross_sales / total_revenues,
log_ratio  = log10(ratio)           # 0 on this scale ⇔ ratio = 1
)

ggplot(ratio_df, aes(x = factor(year), y = log_ratio)) +
geom_violin(fill = "grey85", linewidth = 0.2, alpha = 0.6, trim = TRUE) +
geom_boxplot(width = 0.2, outlier.alpha = 0.3) +
geom_hline(yintercept = 0, linetype = 2) +
scale_y_continuous(
name = "log10(Gross sales / Total revenues)",
breaks = c(-1, -0.5, 0, 0.5, 1),
labels = c("0.1×","0.32×","1×","3.2×","10×")
) +
labs(
title = "How close are gross sales to total revenues?",
x = "Year"
)

```

As expected, most firms' gross sales are near to their total revenues as
the violins sit tightly around 0 on the log scale (≈1×) each year.
Dispersion has a slight right tail and is relatively constant between
years.

# Conclusions

Changes in coverage are unlikely to be the driving force behind patterns
as Sweden's business sample and sector mix remained consistent from 2017
to 2021. With only slight year-to-year variation, balance-sheet size
(assets and liabilities) showed a gentle rising trend, suggesting that
businesses mostly preserved their scale and financing ability under
COVID. Flow metrics (gross sales and total revenues) exhibit a typical
decline in 2020 and a distinct comeback in 2021, usually above 2019
medians. IQRs stay broad but stable, which is consistent with
heavy-tailed distributions rather than a midpoint collapse. Throughout
the years, the gross-sales/total-revenues ratio remained consistently
below 1, indicating that the revenue composition did not significantly
change (i.e., there was no significant shift towards non-sales income).

As the size of both profits and losses increased in 2021, earnings had
the strongest pandemic "signature," suggesting a more divided recovery
with distinct winners and poor performers. Comparing headcounts before
and after the epidemic revealed no statistically significant decline,
suggesting worker hoarding, resilience, and/or supporting policy
settings. According to sector opinions, public administration increased
as pandemic-sensitive areas (like lodging and food) moved more, while
major sectors (like manufacturing, professional services, and wholesale)
stayed the same.

The pandemic caused a brief, moderate revenue shock in 2020, a robust
recovery in 2021, increased earnings volatility, and no widespread
downsizing, or stress without systemic harm. The remaining concerns
which should be considered include survivorship effects, big tails, and
zeros/negatives; sector-specific drills and deeper panel models which
would improve attribution.

# Resources

The materials used for this report are:

1.  Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
    Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller
    E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V,
    Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). "Welcome to
    the tidyverse." *Journal of Open Source Software*, *4*(43), 1686.
    [doi:10.21105/joss.01686](https://doi.org/10.21105/joss.01686).

2.  Bureau van Dijk Electronic Publishing. (n.d.). Osiris. Bureau van
    Dijk Electronic Publishing.
    <https://ezproxy.lib.monash.edu.au/login?url=http://osiris.bvdep.com/ip>

3.  Sweden.se. “The Swedish Economy.” Sweden.se, 19 June 2025. Web.
    Accessed 20 Oct. 2025.
    https://www.sweden.se/work-business/business-in-sweden/the-swedish-economy

4.  OECD. “OECD Economic Surveys: Sweden 2025 – Full Report.” OECD
    Publishing, 5 June 2025. Web. Accessed 20 Oct. 2025.
    https://www.oecd.org/en/publications/oecd-economic-surveys-sweden-2025_75e94b2f-en/full-report.html

5.  LinkedIn. “Top Companies — Sweden.” LinkedIn, n.d. Web. Accessed
    \[Access on 20 october 2026\].
    https://www.linkedin.com/hubs/top-companies/se/

6.  World Bank. (2022). Chapter 1: The economic impacts of the COVID-19
    crisis (in World Development Report 2022). Retrieved October 20,
    2025, from
    https://www.worldbank.org/en/publication/wdr2022/brief/chapter-1-introduction-the-economic-impacts-of-the-covid-19-crisis

7.  NordForsk. (2025, August 4). The pandemic: The economy was put in a
    freezer and the crisis didn’t materialise. NordForsk.
    https://www.nordforsk.org/news/pandemic-economy-was-put-freezer-and-crisis-didnt-materialise

8.  Statista. (n.d.). Number of layoffs after the coronavirus outbreak
    by industry in Sweden \[Statistical table\]. Retrieved <date>, from
    https://www.statista.com/statistics/1110586/number-of-layoffs-after-the-coronavirus-outbreak-by-industry-in-sweden/?srsltid=AfmBOoqZPfe6Toecil3_IHAd7HBKkwSdgG5PsI-5CTRtkTR_nnzaypCL

9.  Zhu H. (n.d.). kableExtra: Construct Complex Table with ‘kable’ and
    Pipe Syntax. R package. Retrieved from
    https://CRAN.R-project.org/package=kableExtra.

10. Waring E., Quinn M., McNamara A., et al. (n.d.). skimr: Compact and
    Flexible Summaries of Data Frames. R package. Retrieved from
    https://CRAN.R-project.org/package=skimr.
